Static and Dynamic Bethe-Salpeter Equations in the $T$-Matrix Approximation

### 2021-11-03 Manuscript/Tmatrix.tex
saving work

@@

Contrary to the $GW$ approximation which relies on the (two-point) dynamically-screened Coulomb potential $W$ computed from a particle-hole random-phase approximation (ph-RPA) problem to target charged excitations (\ie, ionization potentials and electron affinities), \cite{Hedin_1965,Aryasetiawan_1998,Onida_2002,Martin_2016,Reining_2017,Golze_2019} here we consider the $GT$ approximation where one employs the (four-point) $T$ matrix obtained from solving the particle-particle RPA (pp-RPA) equations.
The non-Hermitian pp-RPA problem reads \cite{Schuck_Book,vanAggelen_2013,Peng_2013,Scuseria_2013,Yang_2013,Yang_2013b,Yang_2014a}
\begin{equation}
\label{eq:LR-RPA}
	\begin{pmatrix}
		\bA{}{\ppRPA}			&	\bB{}{\ppRPA}	\\
		-\T{(\bB{}{\ppRPA})}	&	-\bC{}{\ppRPA}	\\
	\end{pmatrix}
	\cdot
	\begin{pmatrix}
		\bX{n}{N\pm2}	\\
		\bY{n}{N\pm2}	\\
	\end{pmatrix}
	=
	\Om{n}{N\pm2}
	\begin{pmatrix}
		\bX{n}{N\pm2}	\\
		\bY{n}{N\pm2}	\\
	\end{pmatrix}
\end{equation}
where the elements of the various matrices are defined as
\begin{subequations}
\begin{align}
	A_{ab,cd}^{\ppRPA} & = \delta_{ab} \delta_{cd} (\e{a}{} + \e{b}{}) + \mel{ab}{}{cd}
	\\ 
	B_{ab,ij}^{\ppRPA} & = \mel{ab}{}{ij}
	\\ 
	C_{ij,kl}^{\ppRPA} & = - \delta_{ik} \delta_{jl} (\e{i}{} + \e{j}{}) +\mel{ij}{}{kl}
\end{align}
\end{subequations}
and 
\begin{equation}
	\mel{pq}{}{rs} = \braket{pq}{rs} - \braket{pq}{sr}
\end{equation}
are two-electron integrals in the spinorbital basis, \ie,
\begin{equation}
	\braket{pq}{rs} = \iint \SO{p}(\bx_1) \SO{q}(\bx_1) \frac{1}{\abs{\br_1 - \br_2}} \SO{r}(\bx_2) \SO{s}(\bx_2)  d\bx_1 d\bx_2
\end{equation}
The pp-RPA problem yields, in the absence of instabilities, $V(V-1)/2$ positive eigenvalues $\Om{n}{N+2}$ and $O(O-1)/2$ negative eigenvalues $\Om{n}{N-2}$, which  correspond respectively to double attachments and double detachments.
The pp-RPA correlation energies is given by \cite{Peng_2013,Scuseria_2013}
\begin{equation}
\begin{split}
	\Ec^\ppRPA 
	& = + \sum_n \Om{n}{N+2} - \Tr(\bA{}{\ppRPA}) 
	\\
	& = - \sum_n \Om{n}{N-2} - \Tr(\bC{}{\ppRPA})
\end{split}
\end{equation}

Based on the knowledge of the pp-RPA eigenvalues and eigenvectors, one can then construct the $T$ matrix.
In the spinorbital basis, the correlation part of the $T$ matrix has the following expression
\begin{equation}
\label{eq:T}
	T_{pq,rs}(\omega) 
		= \sum_{n} \frac{\braket*{pq}{\chi_n^{N+2}}\braket*{rs}{\chi_n^{N+2}}}{\omega - \Om{n}{N+2}}
		+ \sum_{n} \frac{\braket*{pq}{\chi_n^{N-2}}\braket*{rs}{\chi_n^{N-2}}}{\omega - \Om{n}{N-2}}
\end{equation}
with
\begin{subequations}
\begin{align}
	\braket*{pq}{\chi_n^{N+2}} & = \sum_{c < d} \mel{pq}{}{cd} X_{cd,n}^{N+2} + \sum_{k < l}  \mel{pq}{}{kl} Y_{kl,n}^{N+2}
	\\
	\braket*{pq}{\chi_n^{N-2}} & = \sum_{c < d} \mel{pq}{}{cd} X_{cd,n}^{N-2} + \sum_{k < l}  \mel{pq}{}{kl} X_{kl,n}^{N-2}
\end{align}
\end{subequations}

Combining Eqs.~\eqref{eq:G} and \eqref{eq:T}, the correlation part of the $T$-matrix self-energy reads \cite{Romaniello_2012,Martin_2016,Zhang_2017,Li_2021b}
\begin{equation}
	\Sigma^{T}_{pq}(\omega)
	= \sum_{in} \frac{\braket*{pi}{\chi_n^{N+2}}\braket*{qi}{\chi_n^{N+2}}}{\omega + \e{i}{} - \Om{n}{N+2}}
	+ \sum_{an} \frac{\braket*{pa}{\chi_n^{N-2}}\braket*{qa}{\chi_n^{N-2}}}{\omega + \e{a}{} - \Om{n}{N-2}}
\end{equation}

Within the (perturbative) one-shot $GT$ scheme (labeled as $G_0T_0$ in the following), the quasiparticle energies are obtained via linearization of the quasiparticle equation, \ie,
\begin{equation}
	\e{p}{G_0T_0} = \e{p}{\HF} + Z_p \Sigma^{T}_{pp}(\e{p}{\HF})
\end{equation}
where we have assumed a HF starting point and
\begin{equation}
	Z_p = \qty[ 1 - \eval{\pdv{\Sigma^{T}_{pq}(\omega)}{\omega}}_{\omega = \e{p}{\HF}} ]^{-1}
\end{equation}
is the renormalization factor.
Other levels of self-consistency can be considered like the partially self-consistent \textit{``eigenvalue''} $GT$ (ev$GT$) or the quasiparticle self-consistent $GT$ (qs$GT$) scheme.

@@

In order to go beyond the static approximation, it is possible to consider a dynamical correction of the BSE (dBSE) within the Tamm-Dancoff approximation: \cite{Rohlfing_2000,Romaniello_2009b,Ma_2009a,Ma_2009b,Zhang_2013,Rebolini_2016,Olevano_2019,Loos_2020h,Authier_2020,Monino_2021}
\begin{equation}
\label{eq:LR-RPA}
	\begin{pmatrix}
		\bA{}{\dBSE}(\omega)	&	\bB{}{\BSE}	\\
		-\bB{}{\BSE}	&	-\bA{}{\dBSE}(-\omega)	\\
	\end{pmatrix}
	\cdot
	\begin{pmatrix}
		\bX{S}{\dBSE}	\\
		\bY{S}{\dBSE}	\\
	\end{pmatrix}
	=
	\Om{S}{\dBSE}
	\begin{pmatrix}
		\bX{S}{\dBSE}	\\
		\bY{S}{\dBSE}	\\
	\end{pmatrix}
\end{equation}
with 
\begin{equation}
	A_{ia,jb}^{\dBSE}(\omega) = \delta_{ij} \delta_{ab} (\e{a}{\GT} + \e{i}{\GT}) + \mel{ib}{}{aj} + \widetilde{T}_{ij,ba}(\omega)
\end{equation}
where, following Strinati, \cite{Strinati_1988} one can derive the following expression for the elements of the dynamical $T$ matrix
\begin{equation}
\label{eq:dT}
\begin{split}
	\widetilde{T}_{pq,rs}(\omega) 
		& = \sum_{n} \frac{\braket*{pq}{\chi_n^{N+2}}\braket*{rs}{\chi_n^{N+2}}}{\omega - \Om{n}{N+2} + (\e{i}{\GT} + \e{j}{\GT})}
		\\
		& + \sum_{n} \frac{\braket*{pq}{\chi_n^{N-2}}\braket*{rs}{\chi_n^{N-2}}}{\omega + \Om{n}{N-2} - (\e{a}{\GT} + \e{b}{\GT})}
\end{split}
\end{equation}
Equation \eqref{eq:dT} highlights the interesting dynamical structure of the $T$ matrix, where, similarly to the dBSE@$GW$ scheme, the 2h2p configurations are downfolded on the 1h1p configurations.
Although one can potentially access double excitations within this formalism (due to the dynamical nature of the equations), here we rely on the perturbative scheme developed in Ref.~\onlinecite{Loos_2020h} in order to access corrected single excitations.

>>>>

The $GW$ approximation \cite{Hedin_1965} is an elegant resummation of all direct ring diagrams from the particle-hole (ph) channel and is justified in the high-density or weakly-correlated regime. 
Within the $GW$ approximation, a key quantity of many-body perturbation theory, the self-energy $\Sigma$, reads
\begin{equation}
	\Sigma^{GW}(1,2) = i G(1,2) W(1,2)
\end{equation}
where $G(1,2)$ is the one-body Green's function and $W(1,2)$ is the dynamically-screened Coulomb potential, while, \eg, $1 \equiv (\sigma_1,\br_1,t_1)$ is a composite coordinate gathering spin, space, and time variables.

Alternatives to $GW$ do exist. For example, the $T$-matrix (or Bethe-Goldstone) approximation, first introduced in nuclear physics and then in condensed matter physics, sums the ladder diagrams from the particle-particle (pp) channel
and is justified in the low-density or strongly-correlated regime.
While $GW$ considers a two-point interaction, the $T$-matrix approximation relies on a more complex (four-point) effective interaction --- the so-called $T$ matrix --- yielding the following self-energy: 
\begin{equation}
	\Sigma^{GT}(1,2) = i \int G(4,3) T(1,3;2,4) d3 d4
\end{equation}
Combining these two channels, which sounds like a natural thing to do, is also possible and has been explored in the Hubbard dimer. \cite{Romaniello_2012}

@@

Contrary to the $GW$ approximation which relies on the (two-point) dynamically-screened Coulomb potential $W$ computed from a ph-random-phase approximation (ph-RPA) problem to target charged excitations (\ie, ionization potentials and electron affinities), \cite{Hedin_1965,Aryasetiawan_1998,Onida_2002,Martin_2016,Reining_2017,Golze_2019} here we consider the $GT$ approximation where one employs the (four-point) $T$ matrix obtained from solving the pp-RPA equations.
The non-Hermitian pp-RPA problem reads \cite{Schuck_Book,vanAggelen_2013,Peng_2013,Scuseria_2013,Yang_2013,Yang_2013b,Yang_2014a}
\begin{equation}
\label{eq:LR-RPA}
	\begin{pmatrix}
		\bA{}{\ppRPA}			&	\bB{}{\ppRPA}	\\
		-\T{(\bB{}{\ppRPA})}	&	-\bC{}{\ppRPA}	\\
	\end{pmatrix}
	\cdot
	\begin{pmatrix}
		\bX{n}{N\pm2}	\\
		\bY{n}{N\pm2}	\\
	\end{pmatrix}
	=
	\Om{n}{N\pm2}
	\begin{pmatrix}
		\bX{n}{N\pm2}	\\
		\bY{n}{N\pm2}	\\
	\end{pmatrix}
\end{equation}
where the elements of the various matrices are defined as
\begin{subequations}
\begin{align}
	A_{ab,cd}^{\ppRPA} & = \delta_{ab} \delta_{cd} (\e{a}{} + \e{b}{}) + \mel{ab}{}{cd}
	\\ 
	B_{ab,ij}^{\ppRPA} & = \mel{ab}{}{ij}
	\\ 
	C_{ij,kl}^{\ppRPA} & = - \delta_{ik} \delta_{jl} (\e{i}{} + \e{j}{}) +\mel{ij}{}{kl}
\end{align}
\end{subequations}
and 
\begin{equation}
	\mel{pq}{}{rs} = \braket{pq}{rs} - \braket{pq}{sr}
\end{equation}
are two-electron integrals in the spinorbital basis, \ie,
\begin{equation}
	\braket{pq}{rs} = \iint \SO{p}(\bx_1) \SO{q}(\bx_1) \frac{1}{\abs{\br_1 - \br_2}} \SO{r}(\bx_2) \SO{s}(\bx_2)  d\bx_1 d\bx_2
\end{equation}
The pp-RPA problem yields, in the absence of instabilities, $V(V-1)/2$ positive eigenvalues $\Om{n}{N+2}$ and $O(O-1)/2$ negative eigenvalues $\Om{n}{N-2}$, which  correspond respectively to double attachments and double detachments.
The pp-RPA correlation energies is given by \cite{Peng_2013,Scuseria_2013}
\begin{equation}
\begin{split}
	\Ec^\ppRPA 
	& = + \sum_n \Om{n}{N+2} - \Tr(\bA{}{\ppRPA}) 
	\\
	& = - \sum_n \Om{n}{N-2} - \Tr(\bC{}{\ppRPA})
\end{split}
\end{equation}

Based on the knowledge of the pp-RPA eigenvalues and eigenvectors, one can then construct the $T$ matrix.
In the spinorbital basis, the correlation part of the $T$ matrix has the following expression
\begin{equation}
\label{eq:T}
	T_{pq,rs}(\omega) 
		= \sum_{n} \frac{\braket*{pq}{\chi_n^{N+2}}\braket*{rs}{\chi_n^{N+2}}}{\omega - \Om{n}{N+2} + i\eta}
		- \sum_{n} \frac{\braket*{pq}{\chi_n^{N-2}}\braket*{rs}{\chi_n^{N-2}}}{\omega - \Om{n}{N-2} - i\eta}
\end{equation}
with
\begin{subequations}
\begin{align}
	\braket*{pq}{\chi_n^{N+2}} & = \sum_{c < d} \mel{pq}{}{cd} X_{cd,n}^{N+2} + \sum_{k < l}  \mel{pq}{}{kl} Y_{kl,n}^{N+2}
	\\
	\braket*{pq}{\chi_n^{N-2}} & = \sum_{c < d} \mel{pq}{}{cd} X_{cd,n}^{N-2} + \sum_{k < l}  \mel{pq}{}{kl} X_{kl,n}^{N-2}
\end{align}
\end{subequations}

Combining Eqs.~\eqref{eq:G} and \eqref{eq:T}, the correlation part of the $T$-matrix self-energy reads \cite{Romaniello_2012,Martin_2016,Zhang_2017,Li_2021b}
\begin{equation}
	\Sigma^{GT}_{pq}(\omega)
	= \sum_{in} \frac{\braket*{pi}{\chi_n^{N+2}}\braket*{qi}{\chi_n^{N+2}}}{\omega + \e{i}{} - \Om{n}{N+2} + i\eta}
	+ \sum_{an} \frac{\braket*{pa}{\chi_n^{N-2}}\braket*{qa}{\chi_n^{N-2}}}{\omega + \e{a}{} - \Om{n}{N-2} - i\eta}
\end{equation}

Within the (perturbative) one-shot $GT$ scheme (labeled as $G_0T_0$ in the following), the quasiparticle energies are obtained via linearization of the quasiparticle equation, \ie,
\begin{equation}
	\e{p}{G_0T_0} = \e{p}{\HF} + Z_p \Sigma^{GT}_{pp}(\e{p}{\HF})
\end{equation}
where we have assumed a HF starting point and
\begin{equation}
	Z_p = \qty[ 1 - \eval{\pdv{\Sigma^{T}_{pq}(\omega)}{\omega}}_{\omega = \e{p}{\HF}} ]^{-1}
\end{equation}
is the renormalization factor.
Other levels of self-consistency can be considered like the partially self-consistent \textit{``eigenvalue''} $GT$ (ev$GT$) or the quasiparticle self-consistent $GT$ (qs$GT$) scheme.

@@

In order to go beyond the static approximation, it is possible to consider the dynamical version of the BSE (dBSE). \cite{Strinati_1988}
In this case, one must solve the (non-linear) dynamical eigenvalue problem
\begin{equation}
\label{eq:LR-RPA}
	\begin{pmatrix}
		\bA{}{\dBSE}(\Om{S}{\dBSE})	&	\bB{}{\BSE}	\\
		-\bB{}{\BSE}	&	-\bA{}{\dBSE}(-\Om{S}{\dBSE})	\\
	\end{pmatrix}
	\cdot
	\begin{pmatrix}
		\bX{S}{\dBSE}	\\
		\bY{S}{\dBSE}	\\
	\end{pmatrix}
	=
	\Om{S}{\dBSE}
	\begin{pmatrix}
		\bX{S}{\dBSE}	\\
		\bY{S}{\dBSE}	\\
	\end{pmatrix}
\end{equation}
with 
\begin{equation}
	A_{ia,jb}^{\dBSE}(\omega) = \delta_{ij} \delta_{ab} (\e{a}{\GT} + \e{i}{\GT}) + \mel{ib}{}{aj} + \widetilde{T}_{ij,ba}(\omega)
\end{equation}
where, following Strinati, \cite{Strinati_1988} one can derive the following expression for the elements of the dynamical $T$ matrix
\begin{equation}
\label{eq:dT}
\begin{split}
	\widetilde{T}_{pq,rs}(\omega) 
		& = \sum_{n} \frac{\braket*{pq}{\chi_n^{N+2}}\braket*{rs}{\chi_n^{N+2}}}{\omega - \Om{n}{N+2} + (\e{i}{\GT} + \e{j}{\GT}) + i\eta}
		\\
		& + \sum_{n} \frac{\braket*{pq}{\chi_n^{N-2}}\braket*{rs}{\chi_n^{N-2}}}{\omega + \Om{n}{N-2} - (\e{a}{\GT} + \e{b}{\GT}) + i\eta}
\end{split}
\end{equation}
Equation \eqref{eq:dT} highlights the interesting dynamical structure of the $T$ matrix, where, similarly to the dBSE@$GW$ scheme, the 2h2p configurations are downfolded on the 1h1p configurations.
Because solving a non-linear eigenvalue problem is computationally challenging, here we rely on the dynamical perturbative scheme within the Tamm-Dancoff approximation.
\cite{Rohlfing_2000,Romaniello_2009b,Ma_2009a,Ma_2009b,Zhang_2013,Rebolini_2016,Olevano_2019,Loos_2020h,Authier_2020,Monino_2021}
%lthough one can potentially access double excitations within this formalism (due to the dynamical nature of the equations), here we rely on the perturbative scheme developed in Ref.~\onlinecite{Loos_2020h} in order to access corrected single excitations.

The present formalism has been implemented in QuAcK \cite{QuAcK} and we consider here only systems with closed-shell singlet ground states.
The $GW$ and $GT$ calculations are performed by considering a (restricted) HF starting point and standard gaussian basis sets (defined with cartesian functions) are considered.
